---
title: "Final Project Predictive Analytics"
author: "Josh Garzaniti"
date: "2024-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NEW CODE:
```{r}



library(nflverse)
library(nflreadr)

contracts = load_contracts()
  
contracts = contracts %>%
  filter(draft_round <= 7) %>% 
  filter(draft_year >= 1990)


# we can switch the years after draft that we want
sign_year_after_draft = 5  # Set the number of years after draft for re-signing

# Filter and transform the contracts data
resign = contracts %>%
  filter(year_signed > draft_year) %>%
  group_by(player) %>% 
  slice_min(year_signed) %>% 
  mutate(
    years_after_draft = year_signed - draft_year,  # Calculate how many years after the draft the player signed
    resigned_with_draft_team = ifelse(team == draft_team & years_after_draft <= sign_year_after_draft, 1, 0)  # Check if re-signed with draft team within the specified number of years
  ) %>%
  distinct(player, .keep_all = TRUE)  



clean_resign = resign %>%
  select(player, position, team, draft_team, year_signed, gsis_id, 
         draft_year, draft_round, draft_overall, draft_year, years_after_draft, 
         resigned_with_draft_team) %>% 
  # renaming the id column so we can hopefully merge them back after 
  dplyr::rename(player_id = gsis_id)

# pulling the offensive and defensive stats, we can switch the season range too
offense_stats = load_player_stats(seasons = 2000:2024, stat_type = "offense")

offense_stats = offense_stats %>% mutate(position = ifelse( player_display_name == "Trent Richardson", "RB", position ))
offense_stats = offense_stats %>% mutate(position = ifelse( player_display_name == "Mark Walton", "RB", position ))
offense_stats = offense_stats %>% mutate(position = ifelse( player_display_name == "James Casey", "TE", position ))


total_stat_by_year = offense_stats %>%
  group_by(player_id) %>%
  dplyr::rename(season_stats = season) %>% 
  summarize(
    position = first(position),
    
    # sum of the passing stats that are in the list
    completions = sum(completions, na.rm = TRUE),
    attempts = sum(attempts, na.rm = TRUE),
    passing_yards = sum(passing_yards, na.rm = TRUE),
    passing_tds = sum(passing_tds, na.rm = TRUE),
    interceptions = sum(interceptions, na.rm = TRUE),
    sacks = sum(sacks, na.rm = TRUE),
    sack_yards = sum(sack_yards, na.rm = TRUE),
    sack_fumbles = sum(sack_fumbles, na.rm = TRUE),
    sack_fumbles_lost = sum(sack_fumbles_lost, na.rm = TRUE),
    passing_air_yards = sum(passing_air_yards, na.rm = TRUE),
    passing_yards_after_catch = sum(passing_yards_after_catch, na.rm = TRUE),
    passing_first_downs = sum(passing_first_downs, na.rm = TRUE),
    passing_epa = sum(passing_epa, na.rm = TRUE),
    passing_2pt_conversions = sum(passing_2pt_conversions, na.rm = TRUE),
    
    # sum of the rushing stats that are in the list
    carries = sum(carries, na.rm = TRUE),
    rushing_yards = sum(rushing_yards, na.rm = TRUE),
    rushing_tds = sum(rushing_tds, na.rm = TRUE),
    rushing_fumbles = sum(rushing_fumbles, na.rm = TRUE),
    rushing_fumbles_lost = sum(rushing_fumbles_lost, na.rm = TRUE),
    rushing_first_downs = sum(rushing_first_downs, na.rm = TRUE),
    rushing_epa = sum(rushing_epa, na.rm = TRUE),
    rushing_2pt_conversions = sum(rushing_2pt_conversions, na.rm = TRUE),
    
    # sum of the recieving stats that are in the list
    receptions = sum(receptions, na.rm = TRUE),
    targets = sum(targets, na.rm = TRUE),
    receiving_yards = sum(receiving_yards, na.rm = TRUE),
    receiving_tds = sum(receiving_tds, na.rm = TRUE),
    receiving_fumbles = sum(receiving_fumbles, na.rm = TRUE),
    receiving_fumbles_lost = sum(receiving_fumbles_lost, na.rm = TRUE),
    receiving_air_yards = sum(receiving_air_yards, na.rm = TRUE),
    receiving_yards_after_catch = sum(receiving_yards_after_catch, na.rm = TRUE),
    receiving_first_downs = sum(receiving_first_downs, na.rm = TRUE),
    receiving_epa = sum(receiving_epa, na.rm = TRUE),
    receiving_2pt_conversions = sum(receiving_2pt_conversions, na.rm = TRUE),
  )


sum_stats = total_stat_by_year %>% 
  group_by(position, player_id) %>% 
  filter(position %in% c("QB", "RB", "TE", "WR"))



contract_and_stats = clean_resign %>%
  full_join(sum_stats, by = "player_id") 


contract_and_stats = contract_and_stats %>% 
  filter(draft_year >= 1990, na.rm = TRUE, 
         draft_year <= 2020, na.rm = TRUE) %>% 
  filter(position.x %in% c("QB", "RB", "TE", "WR")) %>% 
  filter(!is.na(player_id)) %>% 
  select(-position.y)

contract_and_stats = contract_and_stats[complete.cases(contract_and_stats$completions),]

#changing NA draft teams
contract_and_stats$draft_team[is.na(contract_and_stats$draft_team)] = "49ers"

contract_and_stats = contract_and_stats %>% mutate(resigned_with_draft_team = ifelse(team == draft_team & years_after_draft <= sign_year_after_draft, 1, 0))
```


##Change data types to fit RF model we can’t use player, player id, (draft year and year resigned maybe could work as factors)
```{r}

contract_and_stats$position.x = as.factor(contract_and_stats$position.x)
contract_and_stats$team = as.factor(contract_and_stats$team)
contract_and_stats$draft_team = as.factor(contract_and_stats$draft_team)
contract_and_stats$draft_round = as.factor(contract_and_stats$draft_round)
contract_and_stats$draft_overall = as.numeric(contract_and_stats$draft_overall)
contract_and_stats$years_after_draft = as.numeric(contract_and_stats$years_after_draft)
contract_and_stats$resigned_with_draft_team = as.factor(contract_and_stats$resigned_with_draft_team)
contract_and_stats$completions = as.numeric(contract_and_stats$completions)
contract_and_stats$attempts = as.numeric(contract_and_stats$attempts)
contract_and_stats$passing_yards = as.numeric(contract_and_stats$passing_yards)
contract_and_stats$passing_tds = as.numeric(contract_and_stats$passing_tds)
contract_and_stats$interceptions = as.numeric(contract_and_stats$interceptions)
contract_and_stats$sacks = as.numeric(contract_and_stats$sacks)
contract_and_stats$sack_yards = as.numeric(contract_and_stats$sack_yards)
contract_and_stats$sack_fumbles = as.numeric(contract_and_stats$sack_fumbles)
contract_and_stats$sack_fumbles_lost = as.numeric(contract_and_stats$sack_fumbles_lost)
contract_and_stats$rushing_first_downs = as.numeric(contract_and_stats$rushing_first_downs)
contract_and_stats$passing_air_yards = as.numeric(contract_and_stats$passing_air_yards)
contract_and_stats$passing_yards_after_catch = as.numeric(contract_and_stats$passing_yards_after_catch)
contract_and_stats$passing_first_downs = as.numeric(contract_and_stats$passing_first_downs)
contract_and_stats$passing_epa = as.numeric(contract_and_stats$passing_epa)
contract_and_stats$passing_2pt_conversions = as.numeric(contract_and_stats$passing_2pt_conversions)
contract_and_stats$carries = as.numeric(contract_and_stats$carries)
contract_and_stats$rushing_yards = as.numeric(contract_and_stats$rushing_yards)
contract_and_stats$rushing_tds = as.numeric(contract_and_stats$rushing_tds)
contract_and_stats$rushing_fumbles = as.numeric(contract_and_stats$rushing_fumbles)
contract_and_stats$rushing_fumbles_lost = as.numeric(contract_and_stats$rushing_fumbles_lost)
contract_and_stats$rushing_epa = as.numeric(contract_and_stats$rushing_epa)
contract_and_stats$rushing_2pt_conversions = as.numeric(contract_and_stats$rushing_epa)
contract_and_stats$receptions = as.numeric(contract_and_stats$receptions)
contract_and_stats$targets = as.numeric(contract_and_stats$targets)
contract_and_stats$receiving_yards = as.numeric(contract_and_stats$receiving_yards)
contract_and_stats$receiving_tds = as.numeric(contract_and_stats$receiving_tds)
contract_and_stats$receiving_fumbles = as.numeric(contract_and_stats$receiving_fumbles)
contract_and_stats$receiving_fumbles_lost = as.numeric(contract_and_stats$receiving_fumbles_lost)
contract_and_stats$receiving_air_yards = as.numeric(contract_and_stats$receiving_air_yards)
contract_and_stats$receiving_yards_after_catch = as.numeric(contract_and_stats$receiving_yards_after_catch)
contract_and_stats$receiving_first_downs = as.numeric(contract_and_stats$receiving_first_downs)
contract_and_stats$receiving_epa = as.numeric(contract_and_stats$receiving_epa)
contract_and_stats$receiving_2pt_conversions = as.numeric(contract_and_stats$receiving_2pt_conversions)



```



##Data Partitioning and Random Forest Model (We need to format vars first to either be factors or numericals) (we can also tune this later)
```{r}
contract_and_stats_ungrouped = contract_and_stats %>%
  ungroup() %>%    
  dplyr::select(-player, -player_id, -team)  

summary(contract_and_stats_ungrouped)

str(contract_and_stats_ungrouped)

sum(is.na(contract_and_stats$resigned_with_draft_team) | contract_and_stats$resigned_with_draft_team == "")

train_indices = createDataPartition(contract_and_stats_ungrouped$resigned_with_draft_team, p = 0.60, list = FALSE)

train_data = contract_and_stats_ungrouped[train_indices, ]

valid_data = contract_and_stats_ungrouped[-train_indices, ]

train_data
#Check to see if training data has NA values
sum(is.na(train_data))

colSums(is.na(train_data))

head(valid_data,20)

colSums(is.na(train_data))

train_data_clean = na.omit(train_data)

train_data_clean = train_data_clean%>%
  select(position.x, draft_team, year_signed, draft_year, draft_round, draft_overall, years_after_draft, completions, attempts, passing_yards, passing_tds, interceptions, sacks, sack_yards, sack_fumbles, sack_fumbles_lost, passing_air_yards, passing_yards_after_catch, passing_first_downs, passing_epa, passing_2pt_conversions, carries, rushing_yards, rushing_tds, rushing_fumbles, rushing_fumbles_lost, rushing_first_downs, rushing_epa, rushing_2pt_conversions, receptions, targets, receiving_yards, receiving_tds, receiving_fumbles, receiving_fumbles_lost, receiving_air_yards, receiving_yards_after_catch, receiving_first_downs, receiving_epa, receiving_2pt_conversions, resigned_with_draft_team)

rf_draft_mod = randomForest(resigned_with_draft_team ~.,
                     	data = train_data_clean,
                     	ntree = 500,
                     	nodesize = 1,
                     	mtry = 40) 

rf_pred = predict(rf_draft_mod, valid_data, type = "prob")


rf_pred_class = rep("0", nrow(rf_pred))

rf_pred_class[rf_pred[,2] >= 0.5] = "1"

resign_table = table(rf_pred_class, valid_data$resigned_with_draft_team) 

levels(as.factor(rf_pred_class))
levels(as.factor(valid_data$resigned_with_draft_team))

confusionMatrix(resign_table, positive = "1")
```
## Tuning the RF Model for Error Rate
```{r}
oob_error_model_1 = rf_draft_mod$err.rate[,1] 

plot_dat_model_1 = cbind.data.frame(rep(1:length(oob_error_model_1)), oob_error_model_1) 
names(plot_dat_model_1) = c("trees", "oob_error_model_1")

graph_plot_1 = ggplot(plot_dat_model_1, aes(x = trees, y = oob_error_model_1)) + 
  geom_point(alpha = 0.5, color = "blue") + 
  geom_smooth() + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank()) + 
  labs(x = "Number of Trees", title = "Error Rate v Number of Trees",
       y = "Error Rate")

graph_plot_1

##Optimal number of trees is 300
```
## Try increasing node size
```{r}
nodesizerf = c(1, 10, 25, 50, 100, 200, 500, 1000)

param = expand.grid(nodesizerf)

names(param) = ("nodesizerf") 

res_vec = rep(NA, nrow(param))

for(i in 1:nrow(param)){ 
  set.seed(719) 
  RF_model_2 = randomForest(resigned_with_draft_team ~., 
                      data=train_data_clean,
                      mtry = 40, 
                      importance = FALSE,   
                      ntree = 300, 
                      nodesize = param$nodesizerf[i]) 
  res_vec[i] = 1 - RF_model_2$err.rate[nrow(RF_model_2$err.rate),1] 
}

summary(res_vec)

res_db = cbind.data.frame(param, res_vec) 

names(res_db)[2] = "oob_accuracy"

res_db

##Best Node size for our Model is 50
```
## Creating a 3rd RF with those optimized params
```{r}
rf_draft_mod3 = randomForest(resigned_with_draft_team ~.,
                     	data = train_data_clean,
                     	ntree = 300,
                     	nodesize = 50,
                     	mtry = 40) 

rf_pred = predict(rf_draft_mod3, valid_data, type = "prob")


rf_pred_class = rep("0", nrow(rf_pred))

rf_pred_class[rf_pred[,2] >= 0.5] = "1"

resign_table = table(rf_pred_class, valid_data$resigned_with_draft_team) 

levels(as.factor(rf_pred_class))
levels(as.factor(valid_data$resigned_with_draft_team))

confusionMatrix(resign_table, positive = "1")
```

## Using Cross Validation on Samples
```{r}
set.seed(719)


cv_index = sample(1:5, nrow(train_data_clean), replace = TRUE )

cv_acc = rep(NA, 5)

for(i in 1:5){ 
  cv_train = train_data_clean[cv_index != i, c(1:41)] 
  cv_valid = train_data_clean[cv_index == i, c(1:41)]

  RF_model_4 = randomForest(resigned_with_draft_team ~ . - resigned_with_draft_team, 
                data = cv_train, 
                mtry = 40, 
                ntree = 300, 
                nodesize = 50) 
  
  pred_model_4 = predict(RF_model_4, cv_valid)
  
 CF_pred_model_4 = confusionMatrix(pred_model_4, cv_valid$resigned_with_draft_team)
}

CF_pred_model_4

#Our best model so far is this cross validation model, predicting at almost 70%
```



